A Detailed Analysis of Performance Bottlenecks in MRI Pre-Processing

### 2023-09-06 paper/paper.tex
First comments

\newcommand{\TG}[1]{\color{red}\textsc{From Tristan: }#1\color{black}}
\newcommand{\MD}[1]{\color{magenta}\textsc{From Tristan: }#1\color{black}}
\newcommand{\HL}[1]{\hl{#1}}

@@

\author{Mathieu Dugr\'e, Tristan Glatard}

\begin{document}
\maketitle

\begin{abstract}
	% TODO

\end{abstract}

\section{Introduction}
% Intro: Large dataset, long processing time, I/O intensive (derivatives), focus on MRI, no current study, pipelines?
% **Main question:** What are the main bottleneck in MRI pre-processing?
% **Why:** Performance of neuroimaging pipeline is critical (larger study, clinical use). However, limited study on performance profile.
Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain. Three main MRI techniques exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI). While other modalities such as EEG, CT, and PET exist, we focus on MRIs for their broad adoption and non-invasive property. However, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.

@@

\section{Methods}
\subsection{Infrastructure}
For our experiments, we used the \textit{Slashbin cluster} at Concordia University. The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-240.1.1.el8\_lustre.x86\_64}.

\subsection{Data}
To ensure the performance profiles of the applications derived from our work are inclusive of different population, we used data with a wide range of age and equal distribution of sex (biological). Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.

We used the the OpenNeuro ds004513 v.1.0.2 dataset~\cite{ds004513:1.0.2}. The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts. The within-subject cohort has nine participants (mean age=43, std=7; 4 females) with two session: eyes open and eyes closed. The replication cohort has eleven participants (mean age=27, std=5; 6 females) with only the eyes open session. We profiled the application using both cohort on the eyes open session, unless explicitly specified.

@@

First, we profile these commonly accepted workflows: fMRIPrep anatomical only without and without FreeSurfer, fMRIPrep complete workflow, and dMRIPrep. This let us study the coarse grain performance of workflows for anatomical, functional, and diffusion MRI pre-processing.

@@

Second, to estimate an average performance profile, we aggregate the profiling results of each pipeline across all subjects. Combining the Pareto's law and VTune hotspots report, we present the functions accounting for most of the CPU time up to 80\% of the total cpu time of the application. This heuristic allows us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.

@@

\section{Results}
% TODO
% 2-step approach:
% 1. Which pipeline constitute the long execution time in preprocessing?
% 2. For those, which functions take the most time to execute? (hotspots)
% Any key finding: common point between functions, mem bound, fops, etc.

@@

\subsection{Makespan}
% Identify which tools take the longest to execute.
% This will let us decided which pipeline to investigate more.
Figure~\ref{fig:makespan-1thread} shows the mean and standard deviation of the makespan, in seconds, while using 1 thread for one execution of all subjects.
Unexpectedly, we observe that both ANTS~BrainExtration and ANTS~RegistrationSyN are slower when using single precision compared to double precision.
% TODO Stat test between mean for single and double precision.
\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
	\caption{Profiling of the pipelines using 1 thread.}
	\label{fig:makespan-1thread}
\end{figure}

\subsection{Hotspots}
Figure~\ref{fig:hotspots-fmriprep-anat-subworkflow} depicts the hotspot observed from profiling the fMRIPrep anatomical sub-workflows. In each sub-figure, the x-axis shows the function ID grouped by parent module, which are ordered by CPU time of (1) the modules, then (2) the functions within a module. The function ID are found in the supplementary material Section~\ref{sec:supplementary}. The bars show the average CPU time in seconds of a function and the error bars indicate the standard deviation, with the left y-axis showing the scale. The light-green dots represent the functions cumulative percentage makespan of the application, with the right y-axis showing the scale. The color of the bar indicate the percentage of calls, for the function, that were ``stalled due to demand memory loads and stores''\footnote{\href{https://www.intel.com/content/www/us/en/docs/vtune-profiler/user-guide/2023-0/cpu-metrics-reference.html\#MEMORY-BOUND}{Vtune Memory Bound metric}}, with the left color bar showing the scale.

@@

Similarly, figure~\ref{subfig:hotspots-ants-brainExtraction-fp} shows the same two function accounting for the most time, although in flipped order. Calls from function~2 are memory bound $36.30\pm{4.98}$ percent of the time. Compared to the double precision version of ANTS brainExtraction, the third function runtime is significantly larger. It is an ITK function transforming points while calculating the displacement field of the image.

@@

	\caption{Hotspots profiling of fMRIPrep's anatomical sub-workflows.}
	\label{fig:hotspots-fmriprep-anat-subworkflow}
	% TODO: Describe the plot components. Plots are quite complex so this will be necessary for proper understanding.
\end{figure*}

\begin{figure*}[ht!]
	\centering
	% FSL MCFLIRT
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-fsl-mcflirt}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-mcflirt.png}
	\end{subfigure}
	% FSL FLIRT
	\hfill
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-fsl-flirt}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-flirt.png}
	\end{subfigure}

	\caption{Hotspots profiling of fMRIPrep's functional sub-workflows.}
	\label{fig:hotspots-fmriprep-func-subworkflow}
	% TODO: Describe the plot components. Plots are quite complex so this will be necessary for proper understanding.
\end{figure*}

>>>>

\newcommand{\TG}[1]{\color{blue}\textsc{From Tristan: }#1\color{black}}
\newcommand{\MD}[1]{\color{magenta}\textsc{From Mathieu: }#1\color{black}}
\newcommand{\HL}[1]{\hl{#1}}

@@

\author{Mathieu Dugr\'e, Tristan Glatard \TG{I think Yohan should be co-author given his technical input}}

\begin{document}

\maketitle

\begin{abstract}
	% TODO																
\end{abstract}

\section{Introduction}
% Intro: Large dataset, long processing time, I/O intensive (derivatives), focus on MRI, no current study, pipelines?
% **Main question:** What are the main bottleneck in MRI pre-processing?
% **Why:** Performance of neuroimaging pipeline is critical (larger study, clinical use). However, limited study on performance profile.

\TG{Good intro but I think it would deserve a better logic. You could start by explaining that computational bottlenecks 
have been one of the main bottlenecks to the processing of large cohorts in neuroimaging, that the community is constantly searching for ways to process data faster (HPC, cloud, GPUs, DL models), and that your paper takes a step back by analyzing the computational bottlenecks in order to subsequently focus on their optimization. 
}

Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain. Three main MRI techniques exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI). While other modalities such as EEG, CT, and PET exist \TG{define acronyms}, we focus on MRIs for their broad adoption and non-invasive property. However, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.

@@

\TG{you could summarize the method (vtune profiler) in the intro.}

\section{Methods}
\subsection{Infrastructure}
For our experiments, we used the \textit{Slashbin cluster} at Concordia University. The cluster has eight compute nodes, each configured with two 16-core Intel(R) Xeon(R) Gold 6130 CPU @ \SI{2.10}{\giga\hertz}, \SI{250}{\gibi\byte} of RAM, \SI{126}{\gibi\byte} of tmpfs, six \SI{447}{\gibi\byte} SSDs with the XFS file system (no RAID enabled), CentOS~8.1 and Linux kernel \textit{4.18.0-240.1.1.el8\_lustre.x86\_64}. \TG{I would defer this subsection to later in the methods as it's not the most central one}

\subsection{Data}
To ensure the performance profiles of the applications derived from our work are inclusive of different populations, we used data with a wide range of age and equal distribution of sex (biological). Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.

We used the the OpenNeuro ds004513 v.1.0.2 dataset~\cite{ds004513:1.0.2}. The anatomical, functional, and diffusion data from 20 health individuals was acquired from two different cohorts. The within-subject cohort has nine participants (mean age=43~yrs, std=7~yrs; 4 females) with two session: eyes open and eyes closed. The replication cohort has eleven participants (mean age=27~yrs, std=5~yrs; 6 females) with only the eyes open session. We profiled the application using both cohort on the eyes open session, unless explicitly specified.

@@

First, we profile these commonly accepted workflows: fMRIPrep anatomical only without and without FreeSurfer, fMRIPrep complete workflow, and dMRIPrep. This lets us study the coarse grain performance of workflows for anatomical, functional, and diffusion MRI pre-processing.

@@

Second, to estimate an average performance profile, we aggregate the profiling results of each pipeline across all subjects. Combining the Pareto's law and VTune hotspots report \TG{It would be useful to define these terms}, we present the functions accounting for most of the CPU time up to 80\% of the total CPU time of the application \TG{check wording}. This heuristic allows us to focus our efforts on the performance critical sections of a pipeline. We report the mean and standard deviation of the execution time for those functions.

@@

\section{Results}
\TG{Interesting results but presentation needs to be worked out. This section should be structured with sub-headings summarizing the main findings, for instance: "Interpolation is the main bottleneck", etc}

% TODO
% 2-step approach:
% 1. Which pipeline constitute the long execution time in preprocessing?
% 2. For those, which functions take the most time to execute? (hotspots)
% Any key finding: common point between functions, mem bound, fops, etc.

@@

\subsection{Makespan}
% Identify which tools take the longest to execute.
% This will let us decided which pipeline to investigate more.
Figure~\ref{fig:makespan-1thread} shows the mean and standard deviation of the makespan \TG{define makespan}, in seconds, while using 1 thread for one execution of all subjects.
Unexpectedly, we observe that both ANTS~BrainExtration and ANTS~RegistrationSyN are slower when using single precision compared to double precision.
% TODO Stat test between mean for single and double precision.
\begin{figure}[h]
	\centering
	\includegraphics[width=\linewidth]{figures/makespan-1thread.png}
	\caption{Profiling of the pipelines using 1 thread. \TG{Font size in the figure is too small, it should be similar to the caption font size}}
	\label{fig:makespan-1thread}
\end{figure}

\subsection{Hotspots}

Figure~\ref{fig:hotspots-fmriprep-anat-subworkflow} depicts the hotspots observed from profiling the fMRIPrep anatomical sub-workflows. In each sub-figure, the x-axis shows the function ID grouped by parent module, which are ordered by CPU time of (1) the modules, then (2) the functions within a module. The function ID are found in the supplementary material Section~\ref{sec:supplementary}. The bars show the average CPU time in seconds of a function and the error bars indicate the standard deviation, with the left y-axis showing the scale. The light-green dots represent the functions cumulative percentage makespan of the application, with the right y-axis showing the scale. The color of the bar indicate the percentage of calls, for the function, that were ``stalled due to demand memory loads and stores''\footnote{\href{https://www.intel.com/content/www/us/en/docs/vtune-profiler/user-guide/2023-0/cpu-metrics-reference.html\#MEMORY-BOUND}{Vtune Memory Bound metric}}, with the left color bar showing the scale.

\TG{I'd move the graphical explanations to the figure caption so that the text can focus on observations}

@@

Similarly, figure~\ref{subfig:hotspots-ants-brainExtraction-fp} shows the same two functions accounting for the most time, although in flipped order. Calls from function~2 are memory bound $36.30\pm{4.98}$ percent of the time. Compared to the double precision version of ANTS brainExtraction, the third function runtime is significantly larger. It is an ITK function transforming points while calculating the displacement field of the image.

@@

	\caption{Hotspots profiling of fMRIPrep's anatomical sub-workflows.}
	\label{fig:hotspots-fmriprep-anat-subworkflow}

	% TODO: Describe the plot components. Plots are quite complex so this will be necessary for proper understanding.
\end{figure*}

\begin{figure*}[ht!]
	\centering
	% FSL MCFLIRT
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-fsl-mcflirt}
		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-mcflirt.png}
	\end{subfigure}
	% FSL FLIRT
	\hfill
	\begin{subfigure}[t]{0.49\textwidth}
		\caption{}
		\label{subfig:hotspots-fsl-flirt}

		\includegraphics[width=\textwidth]{figures/hotspots-1thread-fsl-flirt.png}
	\end{subfigure}
	\caption{Hotspots profiling of fMRIPrep's functional sub-workflows.}
	\label{fig:hotspots-fmriprep-func-subworkflow}

	% TODO: Describe the plot components. Plots are quite complex so this will be necessary for proper understanding.
\end{figure*}
